\sysfull

### 2018-04-15 paper/background_conflict-20180414-165358.tex

%\begin{figure}
\begin{wrapfigure}{t!}{0.5\textwidth}
	\begin{subfigure}[t]{\linewidth}
		\centering \includegraphics[width=\columnwidth]{figures/crc_example}
		\caption{Simplified C code snippet of a CRC calculation from~\cite{hicks_mibench2_2016}: per-byte message division by a polynomial; \texttt{NV} denotes non-volatile variable declaration.}\label{fig:crc_example}
	\end{subfigure}
	\begin{subfigure}[t]{\linewidth}
		\centering \includegraphics[width=\columnwidth]{figures/crc_example_war}
		\caption{Execution steps of the loop body in the snippet above: non-volatile checkpointing did not guarantee data consistency as data has been manipulated (line 9) with stale reminder (line 3)}\label{fig:crc_example_war}
	\end{subfigure}
	\caption{\textbf{Code example demonstrating checkpointing gives rise to data inconsistency.}}\label{fig:code_demo_incosistency}
\end{wrapfigure}
%\end{figure}

@@

Energy harvesting devices operate using energy extracted from sources such as radio frequency transmissions and solar energy. These devices elide tethered power or a battery, instead collecting energy into a capacitor, operating when sufficient energy accumulates, and upon depleting the energy, turning off and recharging.
%\TODO{We need a figure that shows what intermittent execution looks like.}
%
%Batteryless operation has a number of important advantages, making intermittent computing an important research domain. Supplying power to billions~\cite{gartner_iot} of embedded computers using batteries is not sustainable. The European Commission estimates that more than 160 kilotons of consumer batteries enter the European Union annually~\cite{eu_batteries_2016}. Batteries are an environmental risk, are fragile, are limited in their number of charge/discharge cycles, and may require costly physical maintenance that is difficult or impossible deployed (e.g. in space~\cite{kicksat}). By contrast, super-capacitors are durable, promising millions of charge/discharge cycles~\cite[Sec. I]{ongaro_pwre_2012}. The main limitation in moving to capacitor-based energy storage is that capacitor energy density---and consequently operating discharge time---is orders of magnitude less than a battery. 
%
%Given current technology development, battery-less systems are best suited for very long-term sensing and monitoring where access to recharge is either prohibitive or impossible. These include battery-less image capture and processing~\cite{naderiparizi_rfid_2015}, animal monitoring~\cite{thomas_jbcs_2012} or implantable~\cite{rodriguez_tbcs_2015} and digestible~\cite{nadeau_naturebio_2017} sensors.
%
There are several energy harvesting battery-less hardware platforms. For instance, computational RFIDs---open-source TI MSP430-based~\cite{wolverine} WISP~\cite{wisp5} (with its variants such as WISPCam~\cite{naderiparizi_rfid_2015}, NFC-WISP~\cite{zhao_rfid_2015} or NeuralWISP~\cite{holleman_biocas_2008}), Moo~\cite{moo}, and commercial ones such as~\cite{medusa_farsens_2017}. Other platforms include ambient backscatter tag~\cite{liu_sigcomm_2013,parks_sigcomm_2014} or battery-less phone~\cite{talla_imwut_2017}.

@@

%\textcolor{red}{reference work by Intel~\cite{baghsorkhi_cgo_2018}}

%Energy-harvesting devices execute software according to the {\em intermittent execution model}. Physically, a device charges until a threshold, operates briefly until its energy is depleted, shuts down recharges, and repeats the cycle.
%
Software running on an energy-harvesting device executes {\em intermittently} because energy sources are not always available to harvest and buffer sufficient operational energy. An intermittent execution is composed of operating periods interspersed with power failures~\cite{dino,chain,alpaca,ratchet}. The frequency of failures depends on the size of the device's energy storage buffer: a larger buffer allows longer operating periods. 
%Energy-harvesters provide input power orders of magnitude less than operating power, making recharging negligible during operation.
%Intermittent execution is different from continuous execution. 
A power failure clears volatile  state (registers, stack, and globals) and non-volatile memory (e.g., FRAM) persists. Upon a power failure, control flows to a prior point in the execution: by default, to the beginning of {\tt main()}. Early intermittent systems preserved progress by periodically checkpointing volatile execution context to non-volatile memory~\cite{mementos,quickrecall}, sometimes using hardware support~\cite{mementos,mottola2017harvos,hibernusplusplus,hibernus,idetic}.

@@

Checkpoints of volatile state preserve intermittent progress, but do not ensure data consistency~\cite{dino,chain,ratchet}. Data may become inconsistent if an attempt to execute some computation {\em writes} to a non-volatile variable, then power fails, then a second attempt to re-execute the same computation incorrectly {\em reads} the value written in the first attempt, rather than the variable's original value. The situation occurs when code includes a WAR dependence between operations that manipulate non-volatile variables~\cite{ratchet,dino,alpaca}---see Figure~\ref{fig:code_demo_incosistency} that illustrates how state can become inconsistent in an intermittent execution using the cyclic redundancy check (CRC) code from MIBench2~\cite{hicks_mibench2_2016}.
%Figure~\ref{fig:code_demo_incosistency} illustrates how state can become inconsistent in an intermittent execution using the cyclic redundancy check (CRC) code from MIBench2~\cite{hicks_mibench2_2016}. The code computes the CRC for an $\texttt{nBytes}$ byte message $\texttt{msg}$ with remainder $\texttt{Rmnd}$.
%Despite checkpointing at line 3 in Fig.~\ref{fig:crc_example}, the code may compute  $\texttt{data}$ incorrectly because of the read of $\texttt{Rmnd}$ on line 4 and the write of $\texttt{Rmnd}$ on line 5. Figure~\ref{fig:crc_example_war} shows an intermittent execution. The second iteration writes $\texttt{Rmnd}$, but power fails before the checkpoint. After restarting, line 9 reads the updated value of $\texttt{Rmnd}$, producing an incorrect $\texttt{data}$ value. Checkpoint-based systems risk violating memory consistency~\cite{dino}. 
To maintain consistency, prior systems~\cite{dino,ratchet} {\em version} a subset of non-volatile data with the checkpoint. Task-based systems~\cite{chain,alpaca}, which we focus on, ensure consistency using programmer, compiler, and runtime support.

@@

Task-based execution models~\cite{dino,chain,alpaca} ask the programmer to decompose their program into tasks. A task is a function with no caller containing arbitrary computation, sensing, and communication. A programmer describes task control-flow as a {\em task graph}. Task flow happens at programmer demarcated transition points that may be conditional on program values. 
%Task-based programming abstractions guarantee that tasks execute {\em atomically}, regardless of power failures. 
Task-based runtime systems ensure task-atomic semantics by ensuring that repeated task executions are idempotent. The key to idempotence is ensuring that non-volatile updates made by an interrupted task are never visible to a future task execution.

@@

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|}
		\hline
		Model & Data Copied to/from NVRAM \\
		\hline\hline
		Mementos~\cite{mementos}	& Registers + Stack     \\
		DINO~\cite{dino}	& Registers + WAR variables \\
		Chain~\cite{chain}	& PC + channel data\\
		Alpaca~\cite{alpaca}	& PC + WAR variables \\
		Ratchet~\cite{ratchet}, Clank~\cite{hicks_isca_2017} & Registers (requires NV main memory) \\
		Region Formation~\cite{baghsorkhi_cgo_2018} & \textcolor{red}{XXX} \\
		\hline
	\end{tabular}
	\caption{\textbf{Memory consistency enforcement overheads of intermittent execution models;} \emph{PC}: program counter, \emph{channel data}: variables explicitly task-shared by the programmer, \emph{WAR variables}: variables involved in WAR dependences (\emph{NV}: non-volatile).}
	\label{table:chechpoint_comparison}
\end{table}

\subsubsection{Costs of Task-based Models}
\label{sec:cost_task-based}

Despite their advantages, the task-based programming models have their limitations. Two major ones being: {\em memory overhead} and {\em inflexibility to dynamic energy harvesting conditions}.

@@

To understand these non-volatile memory overheads, for the MSP430FR5969~\cite{msp430datasheet} MCU we measured the energy consumption of volatile (SRAM) and non-volatile (FRAM) memory operations. We used EDB~\cite{edb} to monitor voltage drop on a capacitor powering the device across 1600 read and write accesses to each memory type, randomizing to avoid caching. Figure~\ref{fig:framEnergy} summarizes the data, showing that a non-volatile memory access consumes around 1.5 times the energy of a volatile access in this device. This observation is also consistent with the TI specification~\cite{}. \todo{add citation to TI specification on FRAM energy cost}{Przemek} In an intermittent device, energy cost corresponds to decreased run time between power failures. The cost of non-volatile memory motivates \sys, which virtualizes memory to primarily use SRAM.

@@

%Przemek placeholder
%\textbf{Task-based Models have Syntactic Limitations.} State of the art task-based programming models, Alpaca~\cite{alpaca}, has substantial limitations that prohibit writing generic programs. First, for Alpaca (and Chain~\cite{chain}), accesses to protected variables must be applied \emph{only} by using the variable's name, i.e. reference by variable address is not possible. This limitation implies that existing third-party libraries, such as MCU-tailored FFT library~\cite{}, cannot be implemented to support Alpaca or Chain (not to mention the impossibility of the development of an intermittent-safe \texttt{memcpy}). Furthermore, if a protected variable needs to be accessed in a code path that is not predictable by the compiler, for instance an Interrupt Service Routine, Alpaca cannot protect that access.

@@

Avoiding excessively costly, non-terminating tasks and short, high-overhead tasks is a programming challenge, given fixed hardware (no support from dedicated circuitry to measure the state of energy buffer) with a fixed energy buffering capacity. Buffer sizes from prior work on intermittently-powered computing hardware vary widely from 20\,$\mu $F~\cite{rodriguez_tbcs_2015} to 0.1\,F~\cite{moo}. Sizing tasks complicates {\em porting} code from one device to another. An excessively costly task on a device with a small energy buffer may be a relatively short task on a device with a larger buffer. The key problem is that using existing systems, by defining tasks statically for a fixed energy buffer, the programmer either risks non-termination or poor performance due to high overhead. \sys's execution model based on {\em task coalescing}---merging or dissecting multiple tasks {\em during} execution based on the program execution behavior, and {\em task downscaling}---shortening task boundaries during execution when even the smallest task does not fit the energy buffer, both models described in Section~\ref{sec:task_adaptation}, is motivated by these observations.

%\textbf{Motivating the Design of \sys's Execution Model.} The challenge of \em{efficiently} executing a program under frequent power failures, lies in the fact the the program needs to be energy aware. This challenge becomes even bigger when \em{generality} is a design requirement and such the system must not require a specific hardware circuitry to offer its efficiency. \em{Fllexibility} is key for wide adaptation, therefore, \sys's execution model should not limit the programmer to a subset of the functionality of the underlying software technology. \em{\sys champions an efficient, general, and flexible intermittent execution.}

>>>>

%\begin{figure}
\begin{wrapfigure}{t!}{0.5\textwidth}
	\begin{subfigure}[t]{\linewidth}
		\centering \includegraphics[width=\columnwidth]{figures/crc_example}
		\caption{Simplified C code snippet of a CRC calculation from~\cite{hicks_mibench2_2016}: per-byte message division by a polynomial; \texttt{NV} denotes non-volatile variable declaration.}\label{fig:crc_example}
	\end{subfigure}
	\begin{subfigure}[t]{\linewidth}
		\centering \includegraphics[width=\columnwidth]{figures/crc_example_war}
		\caption{Execution steps of the loop body in the snippet above: non-volatile checkpointing did not guarantee data consistency as data has been manipulated (line 9) with stale reminder (line 3)}\label{fig:crc_example_war}
	\end{subfigure}
	\caption{\textbf{Code example demonstrating effect of write after read on volatile memory checkpointing.}}\label{fig:code_demo_incosistency}
\end{wrapfigure}
%\end{figure}

@@

Energy harvesting devices operate using energy extracted from sources such as radio frequency transmissions and solar energy. These devices elide tethered power or a battery, instead collecting energy into a capacitor, operating when sufficient energy accumulates, and upon depleting the energy, turning off and recharging.
\TODO{We need a figure that shows what intermittent execution looks like.}
%
%Batteryless operation has a number of important advantages, making intermittent computing an important research domain. Supplying power to billions~\cite{gartner_iot} of embedded computers using batteries is not sustainable. The European Commission estimates that more than 160 kilotons of consumer batteries enter the European Union annually~\cite{eu_batteries_2016}. Batteries are an environmental risk, are fragile, are limited in their number of charge/discharge cycles, and may require costly physical maintenance that is difficult or impossible deployed (e.g. in space~\cite{kicksat}). By contrast, super-capacitors are durable, promising millions of charge/discharge cycles~\cite[Sec. I]{ongaro_pwre_2012}. The main limitation in moving to capacitor-based energy storage is that capacitor energy density---and consequently operating discharge time---is orders of magnitude less than a battery. 
%
%Given current technology development, battery-less systems are best suited for very long-term sensing and monitoring where access to recharge is either prohibitive or impossible. These include battery-less image capture and processing~\cite{naderiparizi_rfid_2015}, animal monitoring~\cite{thomas_jbcs_2012} or implantable~\cite{rodriguez_tbcs_2015} and digestible~\cite{nadeau_naturebio_2017} sensors.
%
Many platforms enable intermittent, battery-less energy harvesting-based computation. For instance, computational RFIDs---open-source TI MSP430-based~\cite{wolverine} WISP~\cite{wisp5} (with its variants such as WISPCam~\cite{naderiparizi_rfid_2015}, NFC-WISP~\cite{zhao_rfid_2015} or NeuralWISP~\cite{holleman_biocas_2008}), Moo~\cite{moo}, and commercial ones such as~\cite{medusa_farsens_2017}. Other intermittently-powered platforms include ambient backscatter tag~\cite{liu_sigcomm_2013,parks_sigcomm_2014} or battery-less phone~\cite{talla_imwut_2017}.

@@

\textcolor{red}{reference work by Intel~\cite{baghsorkhi_cgo_2018}}

%Energy-harvesting devices execute software according to the {\em intermittent execution model}. Physically, a device charges until a threshold, operates briefly until its energy is depleted, shuts down recharges, and repeats the cycle.
%
Software running on an energy-harvesting device executes {\em intermittently} because buffered energy is only available \emph{sometimes}. An intermittent execution is composed of operating periods interspersed with power failures~\cite{dino,chain,alpaca,ratchet}. The frequency of failures depends on the size of the device's energy storage buffer: a larger buffer allows longer operating periods. Energy-harvesters provide input power orders of magnitude less than operating power, making recharging negligible during operation.

Intermittent execution is different from continuous execution. A power failure clears volatile  state (registers, stack, and globals) and non-volatile memory (e.g., FRAM) persists. At a failure, control flows to a prior point in the execution: by default, to the beginning of {\tt main()}. Early intermittent systems preserved progress by periodically checkpointing volatile execution context to non-volatile memory~\cite{mementos,quickrecall}, sometimes using hardware support~\cite{mementos,mottola2017harvos,hibernusplusplus,hibernus,idetic}.

@@

Checkpoints of volatile state preserve intermittent progress, but do not ensure data consistency~\cite{dino,chain,ratchet}. Data may become inconsistent if an attempt to execute some computation {\em writes} to a non-volatile variable, then power fails, then a second attempt to re-execute the same computation incorrectly {\em reads} the value written in the first attempt, rather than the variable's original value. The situation occurs when code includes a WAR dependence between operations that manipulate non-volatile variables~\cite{ratchet,dino,alpaca}.

Figure~\ref{fig:code_demo_incosistency} illustrates how state can become inconsistent in an intermittent execution using the cyclic redundancy check (CRC) code from MIBench2~\cite{hicks_mibench2_2016}. The code computes the CRC for an $\texttt{nBytes}$ byte message $\texttt{msg}$ with remainder $\texttt{Rmnd}$.

Despite checkpointing at line 3 in Fig.~\ref{fig:crc_example}, the code may compute  $\texttt{data}$ incorrectly because of the read of $\texttt{Rmnd}$ on line 4 and the write of $\texttt{Rmnd}$ on line 5. Figure~\ref{fig:crc_example_war} shows an intermittent execution. The second iteration writes $\texttt{Rmnd}$, but power fails before the checkpoint. After restarting, line 9 reads the updated value of $\texttt{Rmnd}$, producing an incorrect $\texttt{data}$ value. Checkpoint-based systems risk violating memory consistency~\cite{dino}. To maintain consistency, prior systems~\cite{dino,ratchet} {\em version} a subset of non-volatile data with the checkpoint. Task-based systems~\cite{chain,alpaca}, which we focus on, ensure consistency using programmer, compiler, and runtime support.

@@

Task-based execution models~\cite{dino,chain,alpaca} ask the programmer to decompose their program into tasks. A task is a function with no caller containing arbitrary computation, sensing, and communication. A programmer describes task control-flow as a {\em task graph}. Task flow happens at programmer demarcated transition points that may be conditional on program values. Task-based programming abstractions guarantee that tasks execute {\em atomically}, regardless of power failures. Task-based runtime systems ensure task-atomic semantics by ensuring that repeated task executions are idempotent. The key to idempotence is ensuring that non-volatile updates made by an interrupted task are never visible to a future task execution.

@@

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|}
		\hline
		Model & Data Copied to/from NVRAM \\
		\hline\hline
		Mementos~\cite{mementos}                             & Reg. + Stack     \\
		DINO~\cite{dino}                                     & Reg. + WAR variables \\
		Chain~\cite{chain}                                   & PC   + Channel data\\
		Alpaca~\cite{alpaca}                                 & PC   + WAR variables \\
		Ratchet~\cite{ratchet}, Clank~\cite{hicks_isca_2017} & Reg. (requires NV main memory) \\
		Region Formation~\cite{baghsorkhi_cgo_2018} & \textcolor{red}{XXX} \\
		\hline
	\end{tabular}
	\caption{\textbf{Memory consistency enforcement overheads of intermittent execution models;} \emph{Reg.}: the entire register file, \emph{PC}: program counter, \emph{Channel data}: variables explicitly task-shared by the programmer, \emph{WAR variables}: variables involved in WAR dependences (\emph{NV}: non-volatile).}
	\label{table:chechpoint_comparison}
\end{table}

\subsubsection{Costs of Task-based Models}

%Task-based models incur costs that motivate \sys.

@@

%\begin{figure}
%\centering
	%\subfloat[Energy cost of \emph{write} operation]{\includegraphics[width=0.49\columnwidth]{figures/fram_write.eps}}
	%\subfloat[Energy cost of \emph{read} operation]{\includegraphics[width=0.49\columnwidth]{figures/fram_read.eps}}
	%\caption{Cost of accessing volatile (SRAM)/non-volatile (FRAM) memory during write (left) and read (right) operation.}
	%\label{fig:framEnergy}
%\end{figure}

To understand these non-volatile memory overheads, for the MSP430FR5969~\cite{msp430datasheet} MCU we measured the energy consumption of volatile (SRAM) and non-volatile (FRAM) memory operations. We used EDB~\cite{edb} to monitor voltage drop on a capacitor powering the device across 1600 read and write accesses to each memory type, randomizing to avoid caching. Figure~\ref{fig:framEnergy} summarizes the data, showing that a non-volatile memory access consumes around 1.5 times the energy of a volatile access in this device. In an intermittent device, energy cost corresponds to decreased run time between power failures. The cost of non-volatile memory motivates \sys, which virtualizes memory to primarily use SRAM.

@@

@@

Avoiding excessively costly, non-terminating tasks and short, high-overhead tasks is a programming challenge, given fixed hardware with a fixed energy buffering capacity. Buffer sizes from prior work vary widely from 20\,$\mu $F~\cite{rodriguez_tbcs_2015} to 0.1\,F~\cite{moo}. Sizing tasks complicates {\em porting} code from one device to another. An excessively costly task on a device with a small energy buffer may be a relatively short, task on a device with a larger buffer. The key problem is that using existing systems, by defining tasks statically for a fixed energy buffer, the programmer either risks non-termination or poor performance due to high overhead. \sys's {\em task coalescing}, described in Section~\ref{sec:task_coalescing}, is motivated by these observations.

\subsection {Motivating the Design of \sys's Execution Model.} 
%% Coala’s model motivation
The challenge of \emph{efficiently} executing a program under frequent power failures, lies in the fact that the program needs to be energy aware.  This challenge becomes even bigger when 
Hardware support is not allowed to advocate the approach \emph{generality} and to eliminate hardware energy consumption overhead. Furthermore, \emph{flexibility} is a key for wide adaptation, therefore, \sys's execution model should not limit a programmer to a subset of the functionalities of the underlying software technology---\emph{\sys champions efficient, general, and flexible intermittent execution.}

%% How does Coala’s execution model progress?
\sys execution model operates on task-based programming model, but It progresses on a dynamic, virtual task granularity. A virtual task can be either multiple static tasks \emph{coalesced} together, when the energy permits, or a sub-task, when it is necessary to preserve forward progress. However, as more tasks coalesce, their  collective commit overhead amortizes better, but the risk of wasting work also increases. If power fails during a long sequence of coalesced tasks, execution will restart from the last commit, i.e. the first task in the sequence, losing  the progress made by any of the coalesced tasks.  The challenge to coalescing tasks is determining how many tasks to coalesce before committing.

%% How does Coala’s execution model decide on its step size when there is abandoned energy?
\sys uses the recent execution history---pure software technique---as a metric to estimate the amount of the available energy and to set the adequate virtual task size accordingly. This fundamental feature prevents \sys from blind virtual task adaptation which can result in a repeated power failures problem. For example, If \sys enlarges the virtual task size by $x$ static tasks on a virtual task completion and reduces it by the same number of static tasks on a power reboot, then \sys may need to reboot, for instance, 10 times to reduce the virtual task size from 10 to one when $x=1$.

%% How does Coala’s execution model decide on its step size when there is scarce energy?

Coala's execution model is neither limited by a static task size to progress, nor by the commit size, i.e. the amount of data that needs to be saved to the non-volatile memory to preserve execution progress. Coala can partition compile-time tasks, if it fails on a single static task more than once, or merge them, if saving the state to non-volatile memory is redundant because a power failure is not expected soon. This unique technique makes the commit size proportional to the number of state-less instructions executed. In other words, if energy allows executing only a small chunk of code, the amount of data to be committed shrinks accordingly.
